
\section*{Preface}

\se\footnote{\sef\ may be read as \textbf{syn}chronous
\textbf{E}mbedded \textbf{R}eactive \textbf{J}ava (with a \textbf{Y}
for the sound).} is a programming language and a design environment for
embedded reactive systems that combines two paradigms:
\begin{itemize}
    \item \emph{Object-oriented modelling} for a robust and flexible
    design.
      
    \item \emph{Synchronous execution} for precise modelling of
    reactive behaviour.
\end{itemize}
Highlights are that
\begin{itemize}
\item \se\ provides a deep embedding of the reactive behaviour into
      the object-oriented data model.
\item \se\ offers fine-grained integration of synchronous formalisms
      such as \esterel~\cite{esterel}, \lustre~\cite{lustre}, and
      \statecharts~\cite{statecharts}.\footnote{We recommend Halbwachs' book
      \emph{Synchronous Programming of Reactive Systems}~\cite{halbwachs} as 	an excellent introduction.}
\end{itemize}

The programming environment supports compilation, configuration,
simulation, and testing, as well as verification by model checking. 
Behavioural descriptions may be edited in graphical or in textual
form.  Code generators for efficient and compact code in C and several
hardware formats are available.

The \se\ language and its programming environment are developed
at the Fraunhofer Institut Autonome Intelligente Systeme.\footnote{The
development of \sef\ has been partially supported by the ESPRIT LTR
SYRF, ``Synchronous formalisms'', and the ESPRIT IIM Project CRISYS,
``Critical Systems and Instrumentation''.}  The programming
environment is freely available with a public domain license.
\newline

\noindent Contact:  
{\tt\small \{reinhard.budde,poigne,sylla\}@ais.fraunhofer.de}.

